Highly Available Transactions: Virtues and Limitations

### 2013-03-23 writeup/hat-vldb.tex
woo; intro pass--sort of crappy
text editing pass

\begin{abstract}
foo bar baz
\end{abstract}

@@

% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
% vldb_sample.bib is the name of the Bibliography in this case
\bibliography{hat-vldb}  
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references

\subsection{References}

Generated by bibtex from your ~.bib file. Run latex, then bibtex, then
latex twice (to resolve references).

@@

\begin{appendix}

\end{appendix}

>>>>

\begin{abstract}
Given the presence of network latency and the possibility of partial
failure and network partitions, many modern distributed systems eschew
transactional functionality, which provides strong semantic guarantees
on groups of multiple operations over multiple data items. In this
work, we consider the problem of providing transactional guarantees
that are highly available, or do not suffer from unavailability during
system partitions or typically incur high network latency.  We
introduce a new taxonomy of highly available algorithms and analyze a
combination of ACID and distributed consistency guarantees to identify
which can and cannot be achieved by Highly Available Transaction (HAT)
systems. In doing so, we analytically and experimentally quantify the
benefit of providing high availability as well as the associated
semantic costs.
\end{abstract}

@@

% The following two commands are all you need in the
% initial runs of your .tex file to
% produce the bibliography for the citations in your paper.
\bibliographystyle{abbrv}
% vldb_sample.bib is the name of the Bibliography in this case
\footnotesize
\bibliography{hat-vldb}  
% You must have a proper ".bib" file
%  and remember to run:
% latex bibtex latex latex
% to resolve all references

@@

%\begin{appendix}

%\end{appendix}

### 2013-03-23 writeup/motivation.tex
text editing pass

\section{Partitions and Latency}

Peter Deutsch begins his list of ``Fallacies of Distributed
Computing'' with two concerns: ``\textit{1.)}  The network is
reliable. \textit{2.)} Latency is zero''~\cite{fallacies-deutsch}. On
a single-node system, communication systems are robust and relatively
fast, and most systems do not need to prepare for partial component
failure. However, as Deutsch points out, these assumptions are no
longer valid in a distributed setting. To extend Jim Gray's storage
latency analogy, if memory is the distance from San Diego to Olympia
(1.5 hours), and disk is as ``far away'' as Pluto (2 years), in a
distributed context, another computer can be even farther---nearing or
even worse than Gray's location of tape access near Andromeda (2,000
years)---with the added concern that messages can be lost in transit,
destroyed by ``black holes'' in the form of network
partitions. Accordingly, accounting for partial failure and
communication expense is core to distributed systems and algorithm
design; ignoring them ``cause[s] big trouble and painful learning
experiences''~\cite{fallacies-deutsch}.

While prominent members of the database community have publicly
speculated about the unimportance of network partitions in distributed
database design~\cite{stonebraker2010errors}, there is mounting
evidence---both anecdotal and experimental---that partitions occur and
latency is non-negligible. In this section, we briefly outline this
evidence and attempt to rigorously quantify the costs of ignoring
network unreliability and latency.

@@

Network partitions---or, informally, communication link failures
between participants in a distributed system---are challenging. A
system facing network partitions faces the impossibility of
simultaneously maintaining a correct ``global'' view of system state
while simultaneously providing operation to all of its clients. The
rate at which a network experiences partitions and the severity and
duration of each partition depends on the given network deployment,
hardware, topology, administration, and a host of other
factors. However, given the \textit{possibility} of partitions, a
system must be prepared to either lose availability of operations or
its ability to reason about global state.

For networks that do not partition, maintaining semantic guarantees
and available operation is
simple~\cite{stonebraker2010errors}. However, this is often not the
case: according to James Hamilton, Vice President and Distinguished
Engineer on the Amazon Web Service team, ``network partitions should
be rare but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence from companies
such as Google~\cite{dean-keynote} confirm Hamilton's statement, while
a host of recent and highly public cloud computing incidents highlight
partition behavior. In April 2011, a network misconfiguration led to a
twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 failure in October 2012 have led
to full site outages for web services like Reddit, Foursquare, and
Heroku~\cite{ec2-downsites}. At a global scale, hardware
failures---like the 2011 outages in Level3's backbones in North
America and Europe due a 2011 bug in Juniper's
routers~\cite{juniper-partition}---and misconfigurations---such as BGP
faults introduced by Pakistan in 2008~\cite{pakistan-youtube} and
academics in 2010~\cite{research-experiment-partition}---can cause
widespread---even Internet-wide---partitions. Many of our discussions
with practitioners---especially those operating on public cloud
infrastructure---confirm that partitions are a reality.

Several recent studies quantify partition behavior. A 2011 study of
several Microsoft datacenters found a mean of 40.8 network link
failures per day (95th percentile: 136), with a median time to repair
of around five minutes, and up to one week. The annual probability of
a device failing and impacting the network is near 10\% for several
critical components like core routers and primary-backup connectors,
while network redundancy only reduces the median impact of failures by
up to 40\%.~\cite{sigcomm-dc}. A 2010 study of five years of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes (95th
percentile at least 34 hours) per link per year~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority WAN, LAN, and configuration-related tickets is ranges
from 114---165 minutes for WAN and LAN issues and up to a day for all
incidents.~\cite{turner2012failure}. The incidence of network failures
is further confirmed by several prior
studies~\cite{labovitz-failures}, showing for example, on Sprint's
WANs, a median time to repair between 2 and 1000 seconds, and a median
time between failures of approximately 3000
seconds.~\cite{ip-backbone-failures}. Indeed, isolating, quantifying,
and designing for network failures is an area of active research in
networking community~\cite{surviving-failures-bodik,
  uw-failure-networks}.

These results---which do not account for server failures (another form
of system partition that has also been quantified at
scale~\cite{google-availability})---indicate that partitions
\textit{do} occur within and across modern datacenters. We do not
intend to fixate on the particular statistical behavior but instead
observe that, as a general trend, partitions exist and must
accordingly be met with either unavailability at some sites or, as we
will discuss, relaxed semantic guarantees.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' Fundamentally,
the speed at which two servers can communicate is (according to modern
physics) bounded by the speed of light. Two servers located on
opposite sides of the Earth face, in the best case---communicating via
a hypothetical link crossing through the center of the
planet---require a minimum 85.1 ms RTT (133.7 ms if sent at ground
level). Of course, not every operation requires contacting another
server around the world, but, as services are replicated to multiple,
geographically distinct sites, the cost of communication grows.

In real deployments, latencies are higher than the speed of light due
to routing, congestion, and processing times; if not, servers a
kilometer apart would enjoy a 6.7 $\mu$s RTT. To illustrate the
difference between intra-datacenter, inter-datacenter, and
inter-planetary networks, we performed a measurement study of network
behavior on Amazon's EC2, a widely-used public compute cloud. We
measured one week of ping times between all seven EC2 regions, across
three ``availability zones'', or closely co-located datacenters, in
the Eastern United States, and within a single ``availabilty zone,''
at a granularity of 1s (dataset will be released).

We summarize the results of our network measurement study in
Table~\ref{table:rtt}. On average, inter-datacenter communication is
between 1.82 and 6.38 times faster than across geographically
co-located datacenters and 40 and 647 times faster than across
geographically distributed datacenters. The cost of wide-area
communication exceeds the speed of light: for example, communicating
from S\~{a}o Paulo to Singapore is lower-bounded by speed-of-light
transit RTT of 106.7 ms but, on average, incurs a 362.8 ms RTT (95th
percentile: 649ms). This is perhaps unsurprising due to modern
Internet topologies, but this three-fold overhead is
non-negligible. As shown in Figure~\ref{fig:rtt}, the distribution of
latencies varies between links, but the trend---in line with other
recent studies~\cite{redblue, mdcc}---is clear: remote communication
has a substantial cost.

@@

>>>>

\section{Motivation}
\label{sec:motivation}

Peter Deutsch begins his classic list of ``Fallacies of Distributed
Computing'' with two concerns fundamental to distributed database
systems: ``\textit{1.)}  The network is reliable. \textit{2.)} Latency
is zero''~\cite{fallacies-deutsch}. On a single server, communication
systems are robust and relatively fast, and most single-node systems
do not need to prepare for partial component failure. However, these
assumptions are no longer valid in a distributed setting. To extend
Jim Gray's storage latency analogy, if memory is 1.5 hours from the
CPU, disk is 2 years away, especially across wide-area networks,
remote server accesses can exceed 30 years, with the added concern
that messages can be lost in transit or delayed indefinitely by
network partitions~\cite{gray-rules}. Accordingly, ignoring network
failures and latency ``cause[s] big trouble and painful learning
experiences''~\cite{fallacies-deutsch}.

In this section, we attempt to quantify the impact of network
partitions and latency in real-world deployments. While the importance
of network behavior is a topic of considerable speculation within the
database community~\cite{stonebraker2010errors}, there is mounting
evidence---both anecdotal and experimental---that partitions occur and
latency is non-negligible.

@@

Network partitions---or, informally, communication link failures
between participants in a distributed system---are challenging. A
system facing network partitions faces the impossibility of
simultaneously maintaining a correct ``global'' view of system state
while simultaneously providing operation to all of its
clients~\cite{davidson-survey}. For networks that do not partition,
maintaining semantic guarantees and available operation is
simple~\cite{stonebraker2010errors}. However, given the
\textit{possibility} of partitions, a system must be prepared to
either lose availability of operations or its ability to reason about
global state.

According to James Hamilton, Vice President and Distinguished Engineer
on the Amazon Web Service team, ``network partitions should be rare
but net gear continues to cause more issues than it
should''~\cite{hamilton-partitions}. Anecdotal evidence confirms
Hamilton's assertion. In April 2011, a network misconfiguration led to
a twelve-hour long series of outages across the Amazon EC2 and RDS
services~\cite{amazon-netpartition}. Subsequent misconfigurations and
partial failures such as another EC2 failure in October 2012 have led
to full site outages for web services like Reddit, Foursquare, and
Heroku~\cite{ec2-downsites}. At a global scale, hardware
failures---like the 2011 outages in Internet backbones in North
America and Europe due a router bug~\cite{juniper-partition}---and
misconfigurations---such as BGP faults in 2008~\cite{pakistan-youtube}
and 2010~\cite{research-experiment-partition}---can cause widespread
partitions. Many of our discussions with practitioners---especially
those operating on public cloud infrastructure---as well as reports
from large-scale operators like Google~\cite{dean-keynote} confirm
that partitions are a reality.

Several recent studies rigorously quantify partition behavior. A 2011
study of several Microsoft datacenters found a mean of 40.8 network
link failures per day (95th percentile: 136), with a median time to
repair of around five minutes, and up to one week. Perhaps
surprisingly, network redundancy only reduces the median impact of
failures by up to 40\%, meaning providers cannot easily reduce the
incidence of partitions~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets.~\cite{turner2012failure}. The incidence of
network failures is further confirmed by several prior studies showing
for example, on Sprint's WANs, a median time to repair between 2 and
1000 seconds, and a median time between failures of approximately 3000
seconds.~\cite{ip-backbone-failures} and frequent path routing
failures on the Internet~\cite{labovitz-failures}. Indeed, isolating,
quantifying, and designing for network failures is an area of active
research in networking community~\cite{surviving-failures-bodik,
  uw-failure-networks}.

These results---which do not consider server-level failures (another
form of system partition that has also been studied at
scale~\cite{google-availability})---indicate that partitions
\textit{do} occur within and across modern datacenters. We do not
intend to fixate on any particular statistical behavior but instead
observe that, as a general trend, partitions exist and must
accordingly be met with either unavailability at some servers or, as
we will discuss, relaxed semantic guarantees.

@@

Even with fault-free networks, distributed systems face the challenge
of communication latency, Deutsch's second ``Fallacy.'' Fundamentally,
the speed at which two servers can communicate is (according to modern
physics) bounded by the speed of light. Two servers located on
opposite sides of the Earth face, in the best case---communicating via
a hypothetical link crossing through the center of the
planet---require a minimum 85.1 ms RTT (133.7 ms if sent at ground
level). As services are replicated to multiple, geographically
distinct sites, the cost of communication increases this upper bound.

In real deployments, latencies are higher than the speed of light due
to routing, congestion, and processing times; if not, servers a
kilometer apart would enjoy a 6.7 $\mu$s RTT. To illustrate the
difference between intra-datacenter, inter-datacenter, and
inter-planetary networks, we performed a measurement study of network
behavior on Amazon's EC2, a widely-used public compute cloud. We
measured one week of ping times between all seven EC2 regions, across
three ``availability zones'', or closely co-located datacenters, in
the Eastern United States, and within a single ``availabilty zone,''
at a granularity of 1s (dataset to be released).

We summarize the results of our network measurement study in
Table~\ref{table:rtt}. On average, inter-datacenter communication (1c)
is between 1.82 and 6.38 times faster than across geographically
co-located datacenters (1b) and 40 and 647 times faster than across
geographically distributed datacenters (1a). The cost of wide-area
communication exceeds the speed of light: for example, communicating
from S\~{a}o Paulo to Singapore is lower-bounded by speed-of-light
transit RTT of 106.7 ms but, on average, incurs a 362.8 ms RTT (95th
percentile: 649ms). As shown in Figure~\ref{fig:rtt}, the distribution
of latencies varies between links, but the trend---in line with other
recent studies~\cite{mdcc, redblue}---is clear: remote communication
has a substantial cost.

@@

\subsection{Modern Systems}
\label{sec:modernacid}

Mitigating the effects of partitions and network latency is a
difficult task, requiring system desigers to make hard trade-offs. In
this section, we briefly discuss known semantic limits for systems
wishing to maintain high availability and their impact on modern
database systems. While the gold standard of ACID databases is
unavailable, we show that many database systems operate under much
weaker models, motivating the study of these models in a highly
available context in the remainder of this paper.

Most notable among existing impossibility results for distributed
systems is Brewer's CAP Theorem, which, as formally proven by Gilbert
and Lynch, states that it is impossible to maintain linearizability
(``C,'' or ``consistency''---albeit poorly named), or the ability to
read the last completed write, and replica availability (``A'') in the
presence of network partitions (``P'')~\cite{gilbert-cap}. Brewer's
Theorem has roots in decades of distributed database
designs~\cite{davidson-survey} and has had a substantial impact on
recent large-scale distributed systems design. Yet, while CAP is
narrowly scoped to the impossibility linearizability, it is often
interpreted to preclude a larger set of guarantees, including ACID, or
transactional atomicity, consistency, isolation and durability
guarantees.

Database designers and researchers have long realized that
serializability---the gold standard of ACID isolation, which
guarantees application-level consistency---is not achievable in a
highly available system~\cite{davidson-survey}.  However, database
systems offer a range of ACID properties besides
serializability. Within a single-node database, the coordination
penalties associated with ensuring equivalence with a serial execution
can be severe and manifest themselves in the form of concurrency
control contention (and, subsequently, performance degradation,
scalability limitations, and, often, external aborts like
deadlocks)~\cite{gray-isolation}. Instead, databases offer a host of
so-called \textit{weak isolation} models that allow varying
restrictions on the space of schedules that are allowable by the
system~\cite{adya}. None of these weak isolation models guarantees
serializability, but the benefits of these models can outweigh the
costs of possible application-level consistency anomalies that arise
from their use.

To understand the prevalence of these weak isolation models, in an
earlier workshop paper~\cite{hat-hotos}, we surveyed the default and
maximum isolation models provided by 18 relational databases, often
claiming to provide ``ACID'' or ``NewSQL'' functionality. As shown in
Table~\ref{table:existing}, only three out of 18 databases provides
serializability by default, and at least eight do not provide
serializability as an option at all. This is particularly surprising
when we consider the widespread deployment of many of these databases,
like Oracle 11g, which are known to power major businesses and product
functionality. Given that these transactional models are frequently
used, our inability to provide serializability in arbitrary HATs is
non-fatal to practical applications. If application writers and
database vendors have already decided that the benefits of weak
isolation outweigh potential application inconsistencies, then, as we
will see, in a highly available environment, similar decisions may be
appropriate.

\begin{table}
\begin{small}
\begin{tabular}{|l|c|c|}
\hline
Database & Default & Maximum\\\hline
Actian Ingres 10.0/10S~\cite{actian-docs} & S & S\\
Aerospike~\cite{aerospike-docs} & RC & RC\\
Akiban Persistit~\cite{akiban-docs} & SI & SI\\
Clustrix CLX 4100~\cite{clustrix-docs} & RR & ?\\
Greenplum 4.1~\cite{greenplum-docs} & RC & S \\
IBM DB2 10 for z/OS~\cite{db2-docs} & CS & S\\
IBM Informix 11.50~\cite{informix-docs} & Depends & RR\\
MySQL 5.6~\cite{innodb-docs} & RR & S \\
MemSQL 1b~\cite{memsql-docs} & RC & RC\\
MS SQL Server 2012~\cite{ms-sql-docs} & RC & S \\
NuoDB~\cite{nuodb-docs} & CR & CR\\
Oracle 11g~\cite{oracle-docs} & RC & SI\\
Oracle Berkeley DB~\cite{bdb-reg-docs} & S & S\\
Oracle Berkeley DB JE~\cite{bdb-je-docs} & RR & S\\
Postgres 9.2.2~\cite{postgres-docs} & RC & S\\
SAP HANA~\cite{hana-docs} & RC & SI\\
ScaleDB 1.02~\cite{scaledb-docs} & RC & RC\\
VoltDB~\cite{voltdb-docs} & S & S\\
\hline
\multicolumn{3}{|p{7cm}|}{{\footnotesize RC: read committed, RR: repeatable read, SI: snapshot isolation, S: serializability, CS: cursor stability, CR: consistent read}}\\\hline

\end{tabular}
\caption{Default and maximum isolation levels for ACID and NewSQL
  databases as of January 2013 (from
  \protect\cite{needed-hotos}).}\vspace{-1.5em}
\label{table:existing}
\end{small}
\end{table}

The primary challenge in providing weak isolation in a HAT context is
that, unfortunately, it is unclear \textit{which} of these isolation
guarantees can be provided with high availability. The existing
algorithms for providing weak isolation are typically presented for a
single-node context and are therefore unavailable. We are not aware of
any prior literature that provides guidance as to the relationship
between weak isolation and high availability. Prior work has examined
the relationship between serializability and high
availability~\cite{davidson-survey} and weak isolation on a
single-server~\cite{adya, ansicritique} but never weak isolation and
high availability taken together. We believe that this apparent gap in
the database literature is a possible contributing factor to the
often-apparent attitude that general-purpose transactional semantics
are either too expensive to provide in many modern distributed
databases---except under special circumstances, such as providing
operations over groups of co-located data items.
